---
title: "Tesing simple hypothesis using the Boltzmann Test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tesing simple hypothesis using the Boltzmann Test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(boltzmannTest)
```
## Introduction
Given is a `data.frame` of $N$ observations and at least one variable. The
variable(s) can be `numeric` (`double` or `integer`) or categorical (`factor`).
The Boltzmann Test tests a (point) hypothesis about the expectation(s) 
$\mathbb{E}[g_i(X)]$ of some functions $g_i(X)$, where $X$ is one of the $N$ 
observations. 

Here, we restrict ourselves to the observed outcomes $\Omega$ rather than
all possible outcomes. With this restriction we can compute expectations
on observed outcomes only
$$
  \mathbb{E}[g(X)] = \sum_{x \in \Omega} p(x)~g(x)~,
$$
where $p(x)$ is the probability for outcome $x \in \Omega$.

We are given the `data.frame` `X`
```{r, exampleData}
X <- data.frame(
  y = c(rep(-1, 40), rep(0, 40), rep(1, 20)),
  s = rep(c("A", "B"), 50),
  stringsAsFactors = TRUE
)
```
with 100 observations for a `numeric` variable called `y` and a `factor` variable
called `s` with labels `A` and `B`.

We can condense the observations of this data set to unique outcomes using 
the function `outcomes_tibble()` 
```{r exampleDataToOutcomes}
(outcomes <- outcomes_tibble(X))
```
which reduces the data `X` from 100 observations to only 6 outcomes. An 
`outcomes_tibble` object is a `tibble` that stores only the unique outcomes 
plus two attributes:

- `empirical` the empirical relative frequencies of an outcome, i.e.
one frequency per row in the `outcomes_tibble`
- `sampleSize` the sample size

It is straightforward to compute the empirical or sample averages for any 
real-valued function of an outcome:
$$
  \mathbb{E}[g(X)] = \sum_{x \in \Omega} f(x)~g(x)~,
$$
where $f(x)$ corresponds to the empirical frequency of outcome $x$.

The Boltzmann Test framework fixes the inference method while leaving a lot
of room for *what to infer* via the choice of the functions $g(X)$. Here, are
some examples (using the data `X`)

- $g(y, s) = 1$ for normalization
- $g(y, s) = y$ for the mean of the numeric variable $y$
- $g(y, s) = \delta(s, \text{A})$ for the rate, i.e. frequency, of the label `A`
in the `factor` variable $s$

But also more complex relationships can be used, e.g. the

- **variance** of $y$
  $$
  g(y, s) = (y - \bar{y})^2
  $$
  with parameter $\bar{y} = \mathbb{E}[g(y, s) = y]$, which can be understood
  as a *structural* expectation: required for the computation of the variance
  but possibly not tested
- **difference of means** of the group $s = \text{B}$ versus $s = \text{A}$
  $$
  g(y, s) = \left[
    \frac{\delta(s, \text{B})}{q} 
    - \frac{\delta(s, \text{A})}{1 - q} 
  \right]~y
  $$
  with parameter $q = \mathbb{E}[g(y_s) =\delta(s, \text{B})]$ as 
  *structural* expectation
  
## One-sample test of the mean



## Two-sample test of the mean
### paired
### unpaired


## Testing the means of more than two groups


## One-sample test of a rate

## Two-sample test of a rate
### paired
### unpaired

## Testing the rates of more than two groups


